Tidal Heating via Internal Gravity Waves in White Dwarf Binaries

### 2019-02-25 writeups/paper.tex
stash

@@

% We study IGW breaking in the plane-parallel approximation in 2D. Known to
% describe both wave steepening (Kloyermeister 1991) and captures mean flow
% absorption (Booker & Bretherton). While the details of breaking are presently
% understood to be 3D (Winters & d'Asaro), we argue in (section) that our
% results should be qualitatively similar to 3D (both 2D and 3D exhibit KHI, and
% since we are able to resolve KHI, the spinup in 3D should be similar?).

In \autoref{s:equations}, we will cover the numerical setup. In \autoref{s:lin}, we
will discuss the agreement of our simulation in the low-amplitude limit with
analytical linear theory. In \autoref{s:nonlin} we will present a few reference
simulations illustrating critical layer absorption depositing horizontal angular
momentum into the fluid.

\section{Equations}\label{s:equations}

We consider an incompressible, isothermal fluid, representative of degenerate
matter in WD bulks, in 2D. We assume a barotropic equation of state as a first
approximation. As we are interested in the behavior of the wave far from the
center of the WD, we approximate the gravitational field as uniform. We model
the background density stratification as $\rho_0(x, z) = \rho_0(z) = \rho_0(z=0)
e^{-z/H}$ for some reference density $\rho_0(z=0) = \at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:fc_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \md{\rho}{t} &= 0 ,\label{eq:fc_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}P}{\rho} + g\hat{z} &= 0.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p*{\vec{u} \cdot \vec{\nabla}}$ is the material
derivative. $\vec{u}, \rho, P$ are the fluid dynamical variables and $-g\hat{z}$
is constant gravitational acceleration. Note that at hydrostatic equilibrium
$\pd{}{t} = 0$ we have $\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0
gH$. A shear flow $u_x(x, z, t) = u_x(z)$ is permitted in hydrostatic
equilibrium, but we will initially consider no shear flow in equilibrium.

\section{Internal Gravity Waves: Linear Theory}

\subsection{Analytical Properties}

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u} \cdot \vec{\nabla}$,
we linearize \autoref{se:fc_orig} by ignoring the advective components of the
material derivative, defining $\rho_1 \equiv \rho - \rho_0 \ll \rho_0 \ll
\rho_0$ and $P_1 \equiv P - P_0 \ll P_0$ deviations from equilibrium and obtain:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\rho_1}{t} - \frac{u_{z} \rho_0}{H} &= 0,\\
        \pd{\vec{u}}{t} + \frac{\vec{\nabla}P_1}{\rho_0}
            + \frac{\rho_1 g\hat{z}}{\rho_0}
            &= 0.
    \end{align}
\end{subequations}

It is well known that the solutions to the $F = 0$ homogeneous
\autoref{se:lin_homo} are of form\cite{drazin,sutherland0}:
\begin{subequations}\label{se:lin_sol}
    \begin{align}
        u_z\p*{x, z, t} &= Ae^{z/2H}e^{i(k_{0x}x + k_{0z}z - \omega_0 t)},\\
        \omega_0^2 &= \frac{N^2k_{0x}^2}{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}},
    \end{align}
\end{subequations}
up to undetermined amplitude $A$, and
\begin{equation}
    N^2 \equiv g^2\p*{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
is the Brunt-V\"ais\"al\"a/buoyancy frequency for our incompressible,
exponentially stratified atmosphere.

@@

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &= \frac{Nk_{0x}}{k_{0x}^2 + k_{0z}^2 + 1/4H^2} \hat{k},\\
            \vec{c}_{g} &= \frac{N\p*{k_{0x}k_{0z}\hat{x}
                - \p*{k_{0z}^2 + \frac{1}{4H^2}}\hat{z}}}
                {\p*{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}}^{3/2}} \nonumber\\
                &\approx \frac{Nk_{0z}}{k_{0x}^2 + k_{0z}^2}\p*{k_{0x}\hat{x} -
                k_{0z}\hat{z}}.
        \end{align}
        We recover the usual result $\vec{c}_{ph} \cdot \vec{c}_g = 0$
        (e.g.\ \cite{drazin,sutherland1}). Note that an IGW transporting energy
        and momentum upwards $c_{g, z} > 0$ has $k_{0z} < 0$.

    \item The time-averaged total $x$-momentum flux in the $\hat{z}$ direction
        can be computed. Since the linear solution is separable as $f(x, z, t) =
        f(z)e^{i(k_{0x}x - \omega_0 t)}$, $x$ averaging and time averaging are
        equivalent, so we may write ($\ev*{\dots}_x$ denotes $x$-averaging)
        \begin{align}
            S_{px, z} &\equiv \ev*{\rho u_x u_z}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_x u_z\;\mathrm{d}x
                    \nonumber\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{0z}}{k_{0x}}.
                    \label{eq:fpx_lin}
        \end{align}

    \item Finally, it may be noted that horizontally averaged horizontal
        momentum $p_x = \ev*{\rho u_x}_x$ obeys $\pd{p_x}{t} + \pd{S_{px}}{z} =
        0$. Since flux is transported on the group velocity
        timescale\cite{sutherland0}, this simplifies to the known expression for
        the wave-induced horizontal mean flow\cite{sutherland0,sutherland1}
        \begin{equation}
            \bar{U}_0 \equiv \ev*{u_x}_x = \frac{\ev*{u_xu_z}_x}{c_{gz}}
        \end{equation}
        We've denoted $c_{gz} = \at{\pd{\omega}{k_{z}}}_{k_{z0}}$ the vertical group
        velocity. This may also be rewritten
        \begin{equation}
            \bar{U}_0 = -\frac{A^2}{2}e^{\frac{z - z_0}{H}}
                \frac{k_{0z}}{k_{0x}}c_{gz}, \label{eq:u0_lin}
        \end{equation}
        where $c_{gz} = -\frac{Nk_{0x}k_{0z}}{\p*{k_{0x}^2 + k_{0z}^2 +
        \frac{1}{4H^2}}^{3/2}}$.
\end{itemize}

@@

To model a continuous IGW wave train excited deep in the WD interior propagating
towards the surface without unnecessary computational cost, we excite an
IGW wave train from the bottom of the simulation domain. We use a volumetric
forcing term to excite IGW at some $z_0 > 0$ interior to the simulation
domain\footnote{Interfacial forcing at the bottom boundary incurs severe CFL
timestepping constraints, as we use a Chebyshev polynomial basis along the $z$
axis in our spectral method which has very small grid spacing at the
boundaries.}. Our forcing excites both IGWs propagating upwards, imitating a
wave tidally excited deeper in the WD, and downwards, which are damped away by
damping layers (see \autoref{ss:numerics}) before inducing strict
Courant-Friedrichs-Lewy (CFL) timestepping constraints.

As not to interfere with the incompressibility constraint, we force the system
on the density equation, replacing \autoref{eq:fc_density} with
\begin{equation}
    \md{\rho}{t} = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p*{k_{0x}x - \omega_0 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying the IGW's dispersion relation for the given
$k_{0x}, \omega_0(k_{0x}, k_{0z})$ will propagate.

The linearized \autoref{eq:vol_drive}, obtained by dropping the advective term
in the material derivative, can be solved analytically up to good accuracy. We
may first approximate the driving term using $e^{\frac{-(z - z_0)^2}{2\sigma^2}}
\approx \sqrt{2\pi \sigma^2}\delta(z - z_0)$ if $\sigma \to 0$\footnote{$\sigma$
must be large enough to be numerically resolved by the spectral code.}. This
system is solved exactly by matching the two homogeneous solutions above and
below $z_0$. We may then approximately relax the solution to $k_{0z}\sigma
\lesssim 1$: an extra factor of $e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises
compared to the $\delta$-function solution (evaluating the Fourier Transform
of $e^{-\frac{(z - z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$), and we obtain
\begin{align}
    u_{1z}&(x, z, t) ={} \frac{Fgk_{0x}^2}{\rho_0(z_0)\omega_0^2}
        \frac{1}{2ik_{0z}}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x + k_{0z}(z - z_0) - \omega_0 t)
            + \frac{1}{2k_{0z}H}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i(k_{0x}x - k_{0z}(z - z_0) - \omega_0 t)
            + \frac{1}{2k_{0z}H}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ moreover is
just $u_{1x} \approx \frac{k_{0z} u_{1z}}{k_{0x}}e^{\frac{1}{2k_{0z}H}}$.

\section{Numerical Simulation}

@@

We nondimensionalize by taking $H = N = \rho_0(z=0) = 1$ in
\autoref{se:fc_orig}.

where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zones. A
Navier-Stokes numerical viscosity $\nu$ is used to damp high wavenumbers and
regularize the nonlinear cascade at near grid resolution: $\nu \sim 0.2
\frac{\omega}{\abs*{k_{0z}}}\frac{L_z}{2\pi N_z}$ was found to be suitable for
$N_z = 1024$, where $k_{0z}$ is the wavenumber of the excited linear mode.
%%%

% Description of spectral method, basis functions, initial conditions
% Grid Size
% Viscosity + diffusivity, damping layers (to simulate radiative BC)
% timestepping? RK443, (CFL limiter??)

The exact implementation of these equations in Dedalus is described in
\autoref{ss:strat_impl}.

\section{Linear Regime}\label{s:lin}

@@

% Nonlinear: Critical layer absorption (viscous = Hazel, nonlin = Winters),
% wave-induced mean flow, horizontal momentum flux

We simulate using Dedalus\cite{dedalus} a spectral numerical method. We use a
Fourier basis in the $x$ direction and a Chebyshev in the $z$ direction. Our
simulation starts with all perturbation quantities at zero, an initial condition
at rest. Our parameters are chosen to be as similar as possible to the later
nonlinear simulations (\autoref{s:nonlin}). We choose as follows:
\begin{itemize}
    \item $k_{0x}$: Astrophysical IGWs have $k_{0x} \ll k_{0z}$, so we choose $k_{0x}
        = \frac{2\pi}{L_x}$ the smallest permitted wavenumber permitted by
        periodic boundary conditions.

    \item We choose $\nu$ $1/10$ that of the nonlinear simulation, to verify
        that the undamped linear solution is recovered.

@@

    \item $S_{px}$ is seen to be constant, as would be expected by
        \autoref{eq:fpx_lin}.

    \item $\bar{U}_0$ agrees with \autoref{eq:u0_lin} at lower $z$ but deviates
        significantly at higher $z$. This deviation is caused by the top damping
        zone that does not conserve horizontal momentum.
\end{itemize}
We also include a qualitatively similar snapshot of the same simulation run
using the $\nu$ from the later nonlinear simulations in \autoref{fig:linear_nu}.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nonu.png}
    \caption{Snapshot of linear simulation after reaching steady state.
        Analytical predictions of $\abs*{u_z}, \bar{U}_0, S_{px}$ are shown in
        orange. Red dotted lines indicate the onset of the damping region and
        blue dotted lines denote the driving region.}\label{fig:linear_nonu}
\end{figure}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nu.png}
    \caption{Same as \autoref{fig:linear_nonu} but with $0.3\times $viscosity
    used for nonlinear simulations (YUBONOTE should be the same but that
    simulation mysteriously stopped without me noticing; restarting worked, so
    currently rerunning). Note the slight resulting dissipation in $S_{px}$.
    }\label{fig:linear_nu}
\end{figure}

\subsection{Nonlinear Regime}\label{s:nonlin}

@@

@@

A purely horizontal shear flow $\bar{U}_0(z) \hat{x}$ can be seen in
\autoref{se:fc_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\bar{U}_0(z)\partial_x$. For a critical value $\omega_0 - \bar{U}_c k_{0x} = 0$,
the frequency of the linear wave in the fluid's frame of reference vanishes and
critical behavior is observed. In a linear theory or a theory where small scales
are viscosity rather than advection dominated, the incident wave has amplitude
reflection and transmission coefficients
\begin{align}
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
    \label{eq:crit_coeffs}
\end{align}
where we have defined Richardson number $\mathrm{Ri} \equiv
\at{\frac{N^2}{\p*{\pd{\bar{U}_0}{z}}^2}}_{z_c}$ at the critical layer $z_c:
\bar{U}_0(z_c) = \frac{\omega_0}{k_{0x}}$. For most shear flows, $\mathrm{Ri}
\gg 1$ and so $\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is
absorbed.

@@

@@

the flux-conservative form of the Euler fluid equations \autoref{se:fc_orig} (we

@@

\autoref{se:fc_var} gives

@@

            - T_1 \pd{\Upsilon}{z},\\

>>>>

Compact white dwarf (WD) binary systems, with orbital periods in the range of
minutes to hours, are important for a range of astrophysical problems. They are
the most important sources of gravitational waves (GWs) for the Laser
Interferometric Space Antenna (LISA)\cite{lisa}. They are also thought to
produce interesting optical transients such as underluminous
supernovae\cite{underlum}, Ca-rich fast transients\cite{carich}, and tidal
novae\cite{tidal_novae}. Most importantly, they have been proposed as the likely
progenitors of type Ia supernovae (e.g.~\cite{Ia0,webbink} or more
recently\cite{Ia1,Ia2}). While presently only a few tens of compact WD binaries
are known\cite{lsst_wd}, \emph{Gaia} (currently gathering data) is expected to
expand the catalog to a few hundreds\cite{lsst_wd} (results based on
\emph{Gaia}'s second data release have already begun to
appear\cite{gaiaDD,gaiaDD2}), and the Large Synoptic Survey Telescope (LSST,
first light scheduled for 2020) will likely detect a few thousand
more\cite{lsst_wd}. These observations will significantly advance the
understanding of WD binaries and their evolution. My proposed theoretical and
computational research is well-timed to take advantage of these new advances.

In spite of the broad importance of WD binaries, the evolution of these systems
prior to their final mergers is not well understood. Much of this uncertainty
comes from our imprecise understanding of tidal interactions, which play an
important role during a compact WD binary's inspiral\cite{fullerII}. Previous
studies have shown that these interactions manifest as tidal excitation of
internal gravity waves (IGW), waves in the WD fluid restored by the buoyancy
force due to density stratification\cite{fullerI}. As these waves propagate
outwards towards the WD surface, they grow in amplitude until they break, as do
ocean waves on a shore, and transfer both energy and angular momentum from the
binary orbit to the outer envelope of the WD\cite{fullerI,fullerII}.

Previous works have found that the dissipation of IGW can generate significantly
more energy than thermal radiation from the isolated WD surface and is thus a
major contributor to the WD energy budget\cite{fullerII,fullerIV}. However,
these works parameterized the wave breaking process in an ad hoc manner. The
details of dissipation, namely the location and spatial extent of the wave
breaking, affect the observable outcome: dissipation near the surface of the WD
can be efficiently radiated away and simply brightens the WD, while dissipation
deep in the WD envelope causes an energy buildup that results in energetic
flares\cite{tidal_novae}. Works in other fields based on numerical simulations
show that strongly nonlinear wave breaking behaves differently than predictions
based in linear and weakly nonlinear theory\cite{winters1994,barker_ogilvie}.
Such fully nonlinear numerical simulations have not been performed for WDs.

@@

% We study IGW breaking in the plane-parallel approximation in 2D. Known to
% describe both wave steepening (klostermeyer 1991) and captures mean flow
% absorption (Booker & Bretherton). While the details of breaking are presently
% understood to be 3D (Winters & d'Asaro), we argue in (section) that our
% results should be qualitatively similar to 3D (both 2D and 3D exhibit KHI, and
% since we are able to resolve KHI, the spinup in 3D should be similar?).

In \autoref{s:equations}, we will describe the system of equations we will
use to analyze IGW breaking. In \autoref{s:theory}, we discuss relevant
analytical results. In \autoref{s:sim} we present the results of numerical
simulations. Finally, in \autoref{s:discussion} we discuss the results of the
preceeding section.

\section{Problem Description}\label{s:equations}

We consider an incompressible, isothermal fluid, representative of degenerate
matter in WD bulks, in 2D. We neglect temperature variations and assume a
barotropic equation of state as a first approximation. As we are interested in
the behavior of the wave far from the center of the WD, we approximate the
gravitational field as uniform. We model the background density stratification
as $\rho_0(x, z) = \rho_0(z) = \rho_0(z=0) e^{-z/H}$ for some reference density
$\rho_0(z=0) = \at{\rho_0(z)}_{z = 0}$.

The Euler equations for an incompressible, barotropic fluid in a uniform
gravitational field are
\begin{subequations}\label{se:nl_orig}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \md{\rho}{t} &= 0 ,\label{eq:nl_density}\\
        \md{\vec{u}}{t} + \frac{\vec{\nabla}p}{\rho} + g\hat{z} &=
            0.\label{eq:nl_mom}.
    \end{align}
\end{subequations}
$\md{}{t} = \pd{}{t} + \p*{\vec{u} \cdot \vec{\nabla}}$ is the Lagrangian or
material derivative. $\vec{u}, \rho, p$ denote the velocity field, density and
pressure respectively. We denote $-g\hat{z}$ constant gravitational
acceleration. Note that at hydrostatic equilibrium $\pd{}{t} = 0$ we have
$\vec{\nabla}P_0 = -\rho_0 g\hat{z}$ and so $P_0 = \rho_0 gH$. A shear flow
$u_x(x, z, t) = u_x(z)$ is permitted in hydrostatic equilibrium, but we will
initially consider no shear flow in equilibrium.

In practice, it is convenient to introduce coordinate $\Upsilon = \ln
\frac{\rho}{\rho_0}$. This both identically enforces $\rho > 0$ and avoids
numerical issues if $\rho$ is small. We also define reduced pressure $P =
\frac{p}{\rho}$. Then, we may rewrite the second two equations of
\autoref{se:nl_orig} as
\begin{subequations}\label{se:nl_upsilon}
    \begin{align}
        \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &= 0,\label{eq:nl_up_density}
            \\
        \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &= 0.
    \end{align}
\end{subequations}
Note that in the new coordinates, hydrostatic equilibrium corresponds to
$\Upsilon_0 = 1, P_0 = gH$.

\section{Internal Gravity Wave Theory}\label{s:theory}

\subsection{Analytical Properties: Linear}

In the small perturbation limit, where flow velocities are small compared to the
characteristic space and time scales $\pd{}{t} \gg \vec{u} \cdot \vec{\nabla}$,
we may linearize \autoref{se:nl_upsilon}. We ignore the advective components of the
material derivative and consider small deviations $\Upsilon_1 = \Upsilon -
\Upsilon_0, P_1 = P - P_0$ about hydrostatic equilibrium:
\begin{subequations}\label{se:lin_homo}
    \begin{align}
        \vec{\nabla} \cdot \vec{u} &= 0,\\
        \pd{\Upsilon_1}{t} - \frac{u_{z}}{H} &= 0,\\
        \pd{\vec{u}}{t} + \vec{\nabla}P_1
            + \p*{1 + \Upsilon_1} g\hat{z}
            &= 0.
    \end{align}
\end{subequations}

This can be solved to obtain solutions up to undetermined amplitude $A$ the
well-known result\cite{drazin,sutherland0}:
\begin{subequations}\label{se:lin_sol}
    \begin{align}
        u_z\p*{x, z, t} &= Ae^{z/2H}e^{i(k_{0x}x + k_{0z}z - \omega_0 t)},\\
        \omega_0^2 &= \frac{N^2k_{0x}^2}{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}},
            \label{eq:disp_rel}
    \end{align}
\end{subequations}
where
\begin{equation}
    N^2 \equiv g^2\p*{\rd{\rho}{P} - \at{\pd{\rho}{P}}_{ad}}
        = \frac{g}{H},
\end{equation}
the Brunt-V\"ais\"al\"a frequency is constant

@@

    \item The phase and group velocities can be computed respectively:
        \begin{align}
            \vec{c}_{ph} &=
                \p*{\frac{\hat{x}}{k_{0x}} + \frac{\hat{z}}{k_{0z}}}\omega,\\
            \vec{c}_{g} &= \frac{N\p*{k_{0x}k_{0z}\hat{x}
                - \p*{k_{0z}^2 + \frac{1}{4H^2}}\hat{z}}}
                {\p*{k_{0x}^2 + k_{0z}^2 + \frac{1}{4H^2}}^{3/2}} \nonumber\\
                &\approx \frac{\omega k_{0z}}{\abs*{\vec{k}_0}^{3/2}}
                    \p*{k_{0x}\hat{x} - k_{0z}\hat{z}}.
        \end{align}
        We recover the usual result $\vec{c}_{ph} \cdot \vec{c}_g = 0$
        (e.g.\ \cite{drazin,sutherland1}). Note that an IGW transporting energy
        and momentum upwards $c_{g, z} > 0$ has $k_{0z} < 0$.

    \item The time-averaged total $x$-momentum flux in the $\hat{z}$ direction
        can be computed. Since the linear solution is separable as $f(x, z, t) =
        f(z)e^{i(k_{0x}x - \omega_0 t)}$, $x$ averaging and time averaging are
        equivalent, so we may write
        \begin{align}
            S_{px, z} &\equiv \ev*{\rho u_x u_z}_x \equiv
                \frac{1}{L_x}\int_0^{L_x}\limits \rho u_x u_z\;\mathrm{d}x
                    \nonumber\\
                &\approx -\frac{A^2}{2}\rho_0(z=0)\frac{k_{0z}}{k_{0x}},
                    \label{eq:fpx_lin}
        \end{align}
        where $\ev*{\dots}_x$ denotes $x$-averaging.
\end{itemize}

@@

To model a continuous IGW wave train excited deep in the WD interior propagating
towards the surface, we use a volumetric forcing term to excite IGW near the
bottom of the simulation domain\footnote{Interfacial forcing at the bottom
boundary incurs strong stability limits, as we use a Chebyshev polynomial basis
along the $z$ axis in our spectral method which has very small grid spacing at
the boundaries.}. Our forcing excites both IGWs propagating upwards, imitating a
wave tidally excited deeper in the WD, and downwards, which are damped away by
the damping layers described in \autoref{ss:damping}.

As not to interfere with the incompressibility constraint, we force the system
on the density equation. This constitutes replacing \autoref{eq:nl_up_density}
with
\begin{equation}
    \md{\Upsilon}{t} + u_z\pd{\ln \rho_0}{z}
        = Fe^{-\frac{(z - z_0)^2}{2\sigma^2}}
            \cos \p*{k_{0x}x - \omega_0 t}.\label{eq:vol_drive}
\end{equation}
Using a narrow Gaussian profile excites a broad $k_z$ wavenumber spectrum, and
only the $k_{0z}$ satisfying dispersion relation \autoref{eq:disp_rel} for the
given $k_{0x}, \omega_0(k_{0x}, k_{0z})$ will propagate.

In the linearized system, the effect of this forcing can be solved analytically
up to good accuracy: we first approximate the driving term using
$e^{\frac{-(z - z_0)^2}{2\sigma^2}} \approx \sqrt{2\pi \sigma^2}\delta(z -
z_0)$, the $\sigma \to 0$ limit\footnote{In practice, $\sigma$ must be large
enough to be numerically resolved by the spectral code.}. This system is solved
exactly by matching the two homogeneous solutions above and below $z_0$. We may
then approximately relax the solution to nonzero $\sigma$: an extra factor of
$e^{-\frac{(k_{0z}\sigma)^2}{2}}$ arises compared to the $\delta$-function
solution (evaluating the Fourier Transform of $e^{-\frac{(z -
z_0)^2}{2\sigma^2}}$ at $k_z = k_{0z}$), and we obtain
\begin{align}
    u_{1z}&(x, z, t) ={} \frac{Fgk_{0x}^2}{\omega_0^2}
        \frac{1}{2ik_{0z}}\frac{e^{\frac{-(k_{0z}\sigma)^2}{2}}}
        {\sqrt{2\pi\sigma^2}} \times\nonumber\\
        &{}\begin{cases}
        e^{\frac{z - z_0}{2H}}e^{i\p*{k_{0x}x + k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z > z_0\\
        e^{\frac{z - z_0}{2H}}e^{i\p*{k_{0x}x - k_{0z}(z - z_0) - \omega_0 t
            + \frac{1}{2k_{0z}H}}}
            & z < z_0\\
    \end{cases}.\label{eq:uz_lin}
\end{align}
The $z > z_0$ solution models an upwards-propagating IGW wavetrain inbound on
the simulation domain from below. The solution for $u_{1x}(x, z, t)$ is
correspondingly $u_{1x} \approx \frac{k_{0z} u_{1z}}{k_{0x}}e^{\frac{i}{2k_{0z}H}}$.

\subsection{Wave Breaking}

In developing the results above, we have neglected advective terms $\pd{}{t} \gg
\vec{u} \cdot \vec{\nabla}$. However, since $\abs*{\vec{u}} \propto e^{z/2H}$,
any nonzero amplitude IGW will eventually violate the linearity criterion. This
regime is relevant in practice since the IGW grows to nonlinear amplitudes
before it reaches the surface of the WD\cite{fullerI,fullerII}. We can coarsely
estimate the height at which this happens by $\abs*{\vec{u}}(z) \gtrsim
\frac{\omega}{\abs*{\vec{k}}}$. This corresponds to when the Lagrangian
displacement $\vec{\xi} \equiv \vec{u} / \omega$ satisfies $\vec{\xi} \cdot
\vec{k} \gtrsim 1$, the same criterion used in\cite{fullerI}.

The general understanding of nonlinear IGW breaking is that the waves'
interactions with the mean flow of the fluid transfer horizontal momentum from
the waves into the fluid; such a process has been conjectured to be responsible
for tidal synchronization in stellar binaries\cite{zahn75,gn89} as well as the
quasi-biennial oscillation\cite{lindzen_qbo}. The details of this process have
been laid out in a few key papers: IGWs are globally unstable to resonant
three-wave interactions\cite{drazin}, causing energy transfer out of the IGW to
daughter modes. These steeper daughter modes facilitate wave breaking,
depositing horizontal momentum in the fluid's mean flow\cite{klostermeyer}. Once
the mean flow reaches the horizontal phase velocity of the parent IGW, a
critical layer forms at which the frequency of the IGW is Doppler-shifted to
zero. The interaction of the parent IGW with the mean flow was first studied in
the inviscid, linear regime in~\cite{booker_bretherton}, which found nearly
complete absorption. This result was reproduced with nonzero
viscosity\cite{hazel}, but weakly nonlinear theory\cite{brown_stewartson} and
fully nonlinear simulations\cite{winters1994} suggest that nonlinear effects can
induce reflection.

An estimate of where these mean flow effects become important can be made using
the wave-induced mean flow formula\cite{eliassen_palm_cite,sutherland0}:
\begin{equation}
     \ev*{u_x} = \bar{U}_0 \equiv \frac{1}{L_x}
        \int\limits u_x\;\mathrm{d}x = \frac{\ev*{u_xu_z}_x}{c_{g,z}},
\end{equation}
where $\ev*{\dots}x$ denotes horizontal averaging. Setting $\bar{U}_0 = c_{ph,
x}$ the critical layer condition, we can solve that $\vec{\xi} \cdot \vec{k}
= \sqrt{\frac{k_x}{k_z}}$ is where the mean flow attains the horizontal phase
velocity. In an astrophysical context, $N \gg \omega$ and $k_z \gg k_x$. Thus,
wave breaking may occur significantly earlier than assumed in current
astrophysics literature.

\section{Internal Gravity Wave Numerical Simulation}\label{s:sim}

To study IGW breaking as a weak perturbation grows to nonlinear amplitudes, we
perform direct numerical simulation using Dedalus\cite{dedalus} a spectral
numerical method. Our region of interest is strictly interior to the WD and so
our simulation must strive to eliminate boundary effects resulting from
nonphysical boundaries. To do so, we adopt periodic boundary conditions in the
$x$ direction and damping layers (described in \autoref{ss:damping}) in the $z$
direction to damp perturbations exiting the domain. Correspondingly, we use a
Fourier basis in the $x$ direction and a Chebyshev in the $z$ direction.

@@

We nondimensionalize by taking $H = N = \rho_0(z=0) = 1$ in
\autoref{se:nl_orig}.

% Grid Size
% Viscosity + diffusivity, damping layers (to simulate radiative BC)
% timestepping? RK443, (CFL limiter??)

Our
simulation starts with all perturbation quantities at zero. We choose as
follows:
\begin{itemize}
    \item $k_{0x}$: Astrophysical IGWs have $k_{0x} \ll k_{0z}$, so we choose $k_{0x}
        = \frac{2\pi}{L_x}$ the smallest permitted wavenumber permitted by
        periodic boundary conditions.

\subsection{Damping Layers}\label{ss:damping}

To imitate an infinite fluid using a finite simulation domain, we use periodic
boundary conditions in the $x$ direction and damping layers at the top and
bottom of the $z$ direction. This damps waves that reach the edge of the
simulation domain without inducing nonphysical reflection. Specifically, these
are implemented by introducing linear damping terms to \autoref{se:nl_upsilon}:
\begin{align}
    \md{\Upsilon}{t} + u_z \pd{\ln \rho_0}{z} &=
        -\Gamma(z)\Upsilon,\nonumber\\
    \md{\vec{u}}{t} + \vec{\nabla}P + P\vec{\nabla}\Upsilon + g\hat{z} &=
        - \Gamma(z)\vec{u},\nonumber\\
    \frac{1}{2\tau}\s*{2 + \tanh \frac{z - z_T}{\Delta z}
        + \tanh \frac{z_B - z}{\Delta z}} &= \Gamma(z),\label{eq:Gamma}
\end{align}
where $z_B = 0.05L_z, z_T = 0.95L_z$ are the boundaries of the damping zone.
This strongly damps perturbations below $z_B$ and above $z_T$ with damping time
$\tau$, negligibly affects dynamics between $z_B, z_T$ and has transition width
governed by $\Delta z$; we use $\Delta z = 0.025L_z$. This prescription, used in
similar studies\cite{lecoanet_damp}, has the advantage of being smooth,
important for spectral methods. As these damping terms are linear, they are
retained in the linearized equations of motion as well.

The full implementation of the fluid equations in Dedalus is described in
\autoref{ss:strat_impl}.

\subsection{Linear Regime}\label{ss:lin}

@@

@@

    \item $S_{px}$ is seen to be constant, as would be expected by
        \autoref{eq:fpx_lin}.
\end{itemize}
We also include a qualitatively similar snapshot of the same simulation run
using the $\nu$ from the later nonlinear simulations in \autoref{fig:linear_nu}.
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nonu.png}
    \caption{Snapshot of linear simulation after reaching steady state.
        Analytical predictions of $\abs*{u_z}, \bar{U}_0, S_{px}$ are shown in
        orange. Red dotted lines indicate the onset of the damping region and
        blue dotted lines denote the driving region.}\label{fig:linear_nonu}
\end{figure}
\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{plots/linear_nu.png}
    \caption{Same as \autoref{fig:linear_nonu} but with $0.3\times $viscosity
    used for nonlinear simulations (YUBONOTE should be the same but that
    simulation mysteriously stopped without me noticing; restarting worked, so
    currently rerunning). Note the slight resulting dissipation in $S_{px}$.
    }\label{fig:linear_nu}
\end{figure}

@@

A
Navier-Stokes numerical viscosity $\nu$ is used to damp high wavenumbers and
regularize the nonlinear cascade at near grid resolution: $\nu \sim 0.2
\frac{\omega}{\abs*{k_{0z}}}\frac{L_z}{2\pi N_z}$ was found to be suitable for
$N_z = 1024$, where $k_{0z}$ is the wavenumber of the excited linear mode.
%%%

@@

A purely horizontal shear flow $\bar{U}_0(z) \hat{x}$ can be seen in
\autoref{se:nl_orig} to have the effect of modifying time derivatives
$\partial_t$ to their frequency in the comoving frame of the fluid $\partial_t -
\bar{U}_0(z)\partial_x$. For a critical value $\omega_0 - \bar{U}_c k_{0x} = 0$,
the frequency of the linear wave in the fluid's frame of reference vanishes and
critical behavior is observed. In a linear theory or a theory where small scales
are viscosity rather than advection dominated, the incident wave has amplitude
reflection and transmission coefficients
\begin{align}
    \mathcal{R} &= e^{-2\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}}, &
    \mathcal{T} &= e^{-\pi \sqrt{\mathrm{Ri} - \frac{1}{4}}},
    \label{eq:crit_coeffs}
\end{align}
where we have defined Richardson number $\mathrm{Ri} \equiv
\at{\frac{N^2}{\p*{\pd{\bar{U}_0}{z}}^2}}_{z_c}$ at the critical layer $z_c:
\bar{U}_0(z_c) = \frac{\omega_0}{k_{0x}}$. For most shear flows, $\mathrm{Ri}
\gg 1$ and so $\mathcal{R}, \mathcal{T} \ll 1$ and the incident wave is
absorbed.

@@

\section{Discussion}\label{s:discussion}

@@

the flux-conservative form of the Euler fluid equations \autoref{se:nl_orig} (we

@@

\autoref{se:nl_var} gives

@@

            - T_1 \pd{\Upsilon}{z}.
